---
title: "Geospatial Visualization with R"
author: "Vedat Erdem Özkul"
institute: "Hochschule Fresenius"

date: "02.06.2026"

subtitle: "------Email: ozkul.vedat@stud.hs-fresenius.de------ Matriculation Number: 401136761"

format:

  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    embed-resources: true

  pdf:
    toc: false
    number-sections: false
    geometry:
      - margin=2.5cm
    fontsize: 11pt

execute:
  echo: true
  warning: false
  message: false
---

## Why Geospatial Visualization Matters

- Understand spatial patterns
- Support decision-making
- Explore relationships between places
- Transform data into visual insight

------------------------------------------------------------------------

## What is Geospatial Data?

Geospatial data is information linked to a location.

Examples:

- Roads
- Buildings
- Cities
- Countries
- Rivers

------------------------------------------------------------------------

## Vector vs Raster Data

| Vector    | Raster              |
|-----------|---------------------|
| Points    | Cells               |
| Lines     | Pixels              |
| Polygons  | Continuous surfaces |
| Roads     | Elevation           |
| Buildings | Temperature         |

------------------------------------------------------------------------

## Layer Concept

A map is built from multiple layers.

Examples:

- Country boundaries
- Cities
- Roads
- Population

Layers can be combined to answer spatial questions.

------------------------------------------------------------------------

## Why Use R?

R provides:

- Reproducible workflows
- Data transformation
- Visualization
- Spatial analysis
- Open-source tools

------------------------------------------------------------------------

## Packages

```{r}
library(sf)
library(tidyverse)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
```

------------------------------------------------------------------------

## Load Spatial Data

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
world <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

germany <- world |>
  filter(admin == "Germany")
```
:::

::: {.column width="50%"}
```{r echo=FALSE}
world <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

germany <- world |>
  filter(admin == "Germany")

germany
```
:::
:::::

------------------------------------------------------------------------

## Coordinate Reference System

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
st_crs(germany)
```
:::

::: {.column width="50%"}
```{r echo=FALSE}
st_crs(germany)
```
:::
:::::

------------------------------------------------------------------------

## Basic Vector Map

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
ggplot() +
  geom_sf(
    data = germany,
    fill = "grey90",
    color = "black"
  ) +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r echo=FALSE}
ggplot() +
  geom_sf(
    data = germany,
    fill = "grey90",
    color = "black"
  ) +
  theme_minimal()
```
:::
:::::

------------------------------------------------------------------------

## A Question You Should Have Asked

Before creating a map:

"How do we know what is inside this dataset?"

We should never use a dataset blindly.

## Useful functions:

names(xxxx)

glimpse(xxxx)

These functions help us discover:

- available variables
- data types
- relevant columns
- possible filters

Only then we can decide what to visualize.

## In-Class Activity: Mapping German Cities

In this activity, we will build a map step by step.

Goal:

- add city points
- combine spatial layers
- filter cities by population
- interpret the result

------------------------------------------------------------------------

## Activity Step 1: Add Cities

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
cities <- ne_download(
  scale = 10,
  type = "populated_places",
  category = "cultural",
  returnclass = "sf"
)

germany_cities <- cities |>
  filter(ADM0NAME == "Germany")
```
:::

::: {.column width="50%"}
```{r echo=FALSE}
cities <- ne_download(
  scale = 10,
  type = "populated_places",
  category = "cultural",
  returnclass = "sf"
)

germany_cities <- cities |>
  filter(ADM0NAME == "Germany")

germany_cities |>
  select(NAME, POP_MAX) |>
  arrange(desc(POP_MAX)) |>
  head(8)
```
:::
:::::

------------------------------------------------------------------------

## Activity Step 2: Combine Spatial Layers

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
ggplot() +
  geom_sf(
    data = germany,
    fill = "grey95",
    color = "black"
  ) +
  geom_sf(
    data = germany_cities,
    size = 0.8,
    alpha = 0.5
  ) +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r echo=FALSE, fig.width=6, fig.height=6}
ggplot() +
  geom_sf(
    data = germany,
    fill = "grey95",
    color = "black"
  ) +
  geom_sf(
    data = germany_cities,
    size = 0.8,
    alpha = 0.5
  ) +
  theme_minimal() +
  labs(
    title = "German Cities",
    subtitle = "Country boundary and populated places"
  )
```
:::
:::::

------------------------------------------------------------------------

## Activity Step 3: Filter the Dataset

::::: columns
::: {.column width="50%"}
```{r eval=FALSE}
germany_cities_major <- germany_cities |>
  filter(POP_MAX > 500000)

ggplot() +
  geom_sf(
    data = germany,
    fill = "grey95",
    color = "black"
  ) +
  geom_sf(
    data = germany_cities_major,
    size = 2
  ) +
  theme_minimal()
```
:::

::: {.column width="50%"}
```{r echo=FALSE, fig.width=6, fig.height=6}
germany_cities_major <- germany_cities |>
  filter(POP_MAX > 500000)

ggplot() +
  geom_sf(
    data = germany,
    fill = "grey95",
    color = "black"
  ) +
  geom_sf(
    data = germany_cities_major,
    size = 2
  ) +
  theme_minimal() +
  labs(
    title = "Major German Cities",
    subtitle = "Cities with population above 500,000"
  )
```
:::
:::::

------------------------------------------------------------------------

## Activity Step 4: Interpret the Map

Discussion questions:

- What changed after filtering?
- Why is filtering useful before visualization?
- Which cities remain visible?
- How can this support decision-making?

## Key Takeaways

- Geospatial data connects information to location.
- Vector and raster data serve different purposes.
- Layers can be combined.
- R provides reproducible geospatial workflows.

------------------------------------------------------------------------

## References

Lovelace, R., Nowosad, J., & Muenchow, J. (2019). *Geocomputation with R*.

Pebesma, E., & Bivand, R. (2023). *Spatial Data Science with Applications in R*.

Huber, S. (n.d.). *How to Use R for Data Science*.

------------------------------------------------------------------------

## Questions?

Thank you.
